Determining Causes of Death from Verbal Autopsy with ChatGPT-4: A Case Study of 6942 Deaths in Sierra Leone from 2019-2022

### 2024-01-21 manuscript/manuscript.tex
Edit abstract

\abstract{\textbf{Background:} Verbal Autopsies (VAs) collect data on deaths and their causes outside of traditional hospital settings to provide more representative counts and Causes of Death (CODs) for reducing premature mortality. Current computer models for COD assignment in VAs perform similar to physicians at the population level, but poorly at the individual level, due to a focus on structured questionnaire data and neglecting free text from the open narratives. Recently, a Generative Pre-trained Transformer (GPT) model called ChatGPT has demonstrated human-level performance on professional and academic exams using free text input. ChatGPT shows promise in mimicking physician behavior for assigning CODs, but to the best of our knowledge, has yet to be tested for assigning CODs in VAs.

\textbf{Methods:} 6942 records collected from VA in Sierra Leone from 2019 to 2022 were used to compare four computer models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for assigning CODs at the population and individual level. All COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for the adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at the population and individual level respectively. Comparisons in CSMF and PCCC among models were evaluated for all records and by COD, age group, and age ranges.

@@

The performances of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, for determining CODs were evaluated using 6939 physician agreed records in 2019 to 2022 from the Health Sierra Leone (HEAL-SL) study \cite{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022}. Physician agreed records are dual-coded records where two physicians had similar COD assignments. COD outputs from the four models were compared to CODs assigned by physicians.

>>>>

\abstract{\textbf{Background:} Verbal Autopsies (VAs) collect data on deaths and their causes outside of traditional hospital settings to provide more representative counts and Causes of Death (CODs) for reducing premature mortality. Current computer models for COD assignment in VAs perform similar to physicians at the population level, but poorly at the individual level, due to a focus on structured questionnaire data and neglecting free text from the open narratives. Recently, a Generative Pre-trained Transformer (GPT) model called ChatGPT has demonstrated human-level performance on professional and academic exams using free text input. ChatGPT shows promise in mimicking physician behavior for assigning CODs, but to the best of our knowledge, has yet to be tested for assigning CODs using open narratives from VAs.

\textbf{Methods:} 6942 records collected from VA in Sierra Leone from 2019 to 2022 were used to compare four computer models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for assigning CODs at the population and individual level. Open narratives were used for Chat-3.5/ChatGPT-4 input, while structured questionnaires were used for InterVA-5/InSilicoVA input. All COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for the adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at the population and individual level respectively. Comparisons in CSMF and PCCC among models were evaluated for all records and by COD, age group, and age ranges.

@@

The performances of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, for determining CODs were evaluated using 6942 physician agreed records in 2019 to 2022 from the Health Sierra Leone (HEAL-SL) study \cite{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022}. Physician agreed records are dual-coded records where two physicians had similar COD assignments. COD outputs from the four models were compared to CODs assigned by physicians.
